---
title: "Differential gene expression analysis"
author: "JG"
date: "`r Sys.Date()`"
output: 
  html_document: default
  github_document: default
editor_options: 
  chunk_output_type: inline
---

## Purpose

- A case study using 30 RNA-seq sample data. 
1. Check overal sample variation using normalized read count
2. Filter for heterogeneity in the data
3. Pair-wise t-test for fold change before and after flowering
4. Cluster progeny and parental samples using reads from selected genes

## Data and links

Base on data in `2024Jun-RNASeq`

## Results and conclusions

1. Overall good sample variation among treatments
2. Identify 19 DGEs that is unique to SG and EAV parental genotypes
3. Cluster result indicates progeny genotype (SGxAV) are more similar to the photoperid sensitive parent (SG)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 12, fig.height = 8)
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
if (!requireNamespace("edgeR", quietly = TRUE)) {
  BiocManager::install("edgeR")
}
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}
pacman::p_load(edgeR, RColorBrewer, rtracklayer, dplyr)
rm(list = ls())
#setwd('~/Downloads')
```

### Import data
```{r, cache = T}
# ------------------- import data --------------- #
DGE_norm <- readRDS('/Users/neoji/Dropbox/Hemp/2024Jun-RNASeq/2024Jun-RNASeq/Gene counts/rRNA-filtered_and_normalized_DGEList.rds')
mdt <- DGE_norm$samples

files <- paste0('/Users/neoji/Dropbox/Hemp/2024Jun-RNASeq/2024Jun-RNASeq/Gene counts/raw gene counts/', dir('/Users/neoji/Dropbox/Hemp/2024Jun-RNASeq/2024Jun-RNASeq/Gene counts/raw gene counts/'))
DGE_raw <- readDGE(files, sep = ' ')

my_obj <- import("/Users/neoji/Dropbox/Hemp/2024Jun-RNASeq/2024Jun-RNASeq/GCF_029168945.1_ASM2916894v1_genomic.gtf.gz")
class(my_obj)
dt_GTF <- data.table::as.data.table(my_obj@elementMetadata@listData)
```

### Check overall sample DGE variation
```{r}
# ------------ Check overall sample variation using plotMDS ------------ #
lcpm <- cpm(DGE_norm, log=TRUE)
par(mfrow=c(1,2))
group <- DGE_norm$samples$group
cultivar <- DGE_norm$samples$cultivar
col.group <- as.factor(cultivar)
levels(col.group) <-  brewer.pal(nlevels(col.group), "Set1")
col.group <- as.character(col.group)
plotMDS(lcpm, labels=cultivar, col=col.group)

sampling_stage <- DGE_norm$samples$growth.stage
col.group <- as.factor(sampling_stage)
levels(col.group) <-  brewer.pal(nlevels(col.group), "Set1")
col.group <- as.character(col.group)
plotMDS(lcpm, labels=sampling_stage, col=col.group)
```


### Filter for outlier data
```{r}
# ------------ filter out extremely skewed data ------------- #
design <- model.matrix(~0+group)
colnames(design) <- gsub("group", "", colnames(design))
colnames(design)

cutoff <- 1
drop <- which(apply(cpm(DGE_norm), 1, max) < cutoff)
DGE_norm_filtered <- DGE_norm[-drop,] 
dim(DGE_norm_filtered) # number of genes left
v0 <- voom(DGE_norm, design, plot=TRUE)
v1 <- voom(DGE_norm_filtered, design, plot=TRUE)
```

### Model pair-wise constrasts
```{r}
# ------------ contrast individual groups ---------- #
v_filter <- voom(DGE_norm_filtered, design, plot=F)
vfit_filtered <- lmFit(v_filter, design)
colnames(coef(vfit_filtered))

# -------- pair-wise contrasts ------- #
design.pairs <- function(levels) {
    n <- length(levels)
    design <- matrix(0,n,choose(n,2))
    rownames(design) <- levels
    colnames(design) <- 1:choose(n,2)
    k <- 0
    for (i in 1:(n-1))
      for (j in (i+1):n) {
        k <- k+1
        design[i,k] <- 1
        design[j,k] <- -1
        colnames(design)[k] <- paste(levels[i],"-",levels[j],sep="")
      }
    design
  }

par(mfrow=c(1,1))
contr.matrix <- design.pairs(colnames(coef(vfit_filtered)))
```


### Detect significant DGEs
```{r}
# using eBayes method to detect DGE
efit <- eBayes(contrasts.fit(vfit_filtered, contrasts=contr.matrix))
plotSA(efit, main="Final model: Mean-variance trend")
# using fold-change t-test method to detect DGE
tfit <- treat(contrasts.fit(vfit_filtered, contrasts=contr.matrix), lfc=1)
# let's test using a strict 'treat' method for now

comb_treats <- colnames(contr.matrix)

ls.top.treat <- list()
for (i in comb_treats) {
  ls.top.treat[[i]] <- topTreat(tfit, coef=i, n=Inf)
}

ls.top.table <- list()
for (i in comb_treats) {
  ls.top.table[[i]] <- topTable(efit, coef=1, sort.by = "P", n = Inf)
}

head(ls.top.table[['EAV-EAV_F']], 5)
sum(ls.top.table[['EAV-EAV_F']]$adj.P.Val < 0.05)
head(ls.top.treat[['EAV-EAV_F']], 5)
sum(ls.top.treat[['EAV-EAV_F']]$logFC > 1)
```

## Select a case to compare
### detect SGxAV progeny variation
```{r}
# ----------- select a pair to find DE gene candidates -------------- #
dt <- decideTests(tfit)
summary(dt)
comb_treats
sel_pair_1 <- which(comb_treats == 'EAV-EAV_F')
sel_pair_2 <- which(comb_treats == 'SG-SG_F')

vennDiagram(dt[,c(sel_pair_1, sel_pair_2)],
            circle.col=c("turquoise", "salmon"))

de.common <- which(dt[,sel_pair_1]!=0 & dt[,sel_pair_2]!=0)
length(de.common)

de.unique <- which((dt[,sel_pair_1]!=0 & dt[,sel_pair_2]==0) | (dt[,sel_pair_1]==0 & dt[,sel_pair_2]!=0))
length(de.unique)

dimnames(dt)[[1]][de.unique][1:10]
```

### Join annotation data
```{r}
# ----- join annotation data ----- #
dt_gene_cand <- data.table::data.table(gene_id = dimnames(dt)[[1]][de.unique])
dt_gene_cand <- dplyr::left_join(dt_gene_cand, dt_GTF, by = 'gene_id')
dt_gene_cand <- filter(dt_gene_cand, gbkey == 'Gene')
```


### Cluster samples involved in the case study
```{r}
# ------- heatmap -------- #
library(gplots)
mycol <- colorpanel(1000,"blue","white","red")
dt_plot <- lcpm[dt_gene_cand$gene_id, ]
colnames(dt_plot) <- paste0(group, paste0('_',seq(1,3)))
dt_plot_f <- dt_plot[, grepl('EAV|SG', colnames(dt_plot))]

heatmap.2(dt_plot_f, scale="row",
          labRow=dt_gene_cand$description, labCol=colnames(dt_plot_f),
          col=mycol, trace="none", 
          margin=c(6,25),
          cexCol=1, dendrogram="column")

# png(paste0('DGE_hm_', comb_treats[sel_pair_1], '_VS_',
#            comb_treats[sel_pair_2], '.png'), plot, width = 14, height = 8, units = 'in', res = 200)
# 
# heatmap.2(dt_plot_f, scale="row",
#           labRow=dt_gene_cand$description, labCol=colnames(dt_plot_f),
#           col=mycol, trace="none", density.info="none",
#           margin=c(12,32), 
#           lhei=c(2,10), 
#           cexRow=1.2,
#           cexCol=1.5,
#           adjRow = c(0,NA),
#           #srtRow=45,
#           dendrogram="column")
# 
# dev.off()

dt_gene_common <- data.table::data.table(gene_id = dimnames(dt)[[1]][de.common])
dt_gene_common <- dplyr::left_join(dt_gene_common, dt_GTF, by = 'gene_id')
dt_gene_common <- filter(dt_gene_common, gbkey == 'Gene')
```

## Case study #2
### Compare LIFTER and EAV
```{r}
sel_pair_1 <- which(comb_treats == 'EAV-EAV_F')
sel_pair_2 <- which(comb_treats == 'Lifter-Lifter_F')

vennDiagram(dt[,c(sel_pair_1, sel_pair_2)],
            circle.col=c("turquoise", "salmon"))

de.common <- which(dt[,sel_pair_1]!=0 & dt[,sel_pair_2]!=0)
length(de.common)

de.unique <- which((dt[,sel_pair_1]!=0 & dt[,sel_pair_2]==0) | (dt[,sel_pair_1]==0 & dt[,sel_pair_2]!=0))
length(de.unique)

dimnames(dt)[[1]][de.unique][1:10]


```


### Cluster samples involved in the case study
```{r}
# ------- heatmap -------- #
dt_gene_cand <- data.table::data.table(gene_id = dimnames(dt)[[1]][de.unique])
dt_gene_cand <- dplyr::left_join(dt_gene_cand, dt_GTF, by = 'gene_id')
dt_gene_cand <- filter(dt_gene_cand, gbkey == 'Gene')

dt_plot <- lcpm[dt_gene_cand$gene_id, ]
colnames(dt_plot) <- paste0(group, paste0('_',seq(1,3)))
mycol <- colorpanel(1000,"blue","white","red")
dt_plot_f <- dt_plot[, grepl('EAV|Lifter', colnames(dt_plot))]

heatmap.2(dt_plot_f, scale="row",
          labRow=dt_gene_cand$description, labCol=colnames(dt_plot_f),
          col=mycol, trace="none", 
          margin=c(6,25),
          cexCol=1, dendrogram="column")
dt_gene_common <- data.table::data.table(gene_id = dimnames(dt)[[1]][de.common])
dt_gene_common <- dplyr::left_join(dt_gene_common, dt_GTF, by = 'gene_id')
dt_gene_common <- filter(dt_gene_common, gbkey == 'Gene')
test <- dt_gene_common


```